# Raster Data Handling {#sec-raster-basics}

```{r chap_4_setup, echo = FALSE, results = "hide", cache = FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,
  cache = FALSE,
  # cache = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  tidy = FALSE,
  cache.lazy = FALSE
)

suppressMessages(library(here))
opts_knit$set(root.dir = here())
```

```{r , eval = FALSE, echo = FALSE}
setwd(here())
```

```{r, include = FALSE, cache = FALSE}
#--- load packages ---#
library(data.table)
library(stringr)
library(raster)
library(terra)
library(lubridate)
library(sf)
library(tictoc)
# setwd("/Users/tmieno2/Box/Teaching/AAEA R/GIS")
```

## Before you start {-}

In this chapter, we will explore how to handle raster data using the `raster` and `terra` packages. The `raster` package has long been the standard for raster data handling, but the `terra` package has now superseded the `raster` package. `terra` typically offers faster performance for many raster operations compared to `raster`. However, we will still cover `raster` object classes and how to convert between `raster` and `terra` objects. This is because many of the existing spatial packages still rely on `raster` object classes and have not yet transitioned to `terra`. Both packages share many function names, and key differences will be clarified as we proceed. 

For many scientists, one of the most common and time-consuming raster data task is extracting values for vector data. As such, we will focus on the essential knowledge needed for this process, which will be thoroughly discussed in @sec-int-RV. 

Finally, if you frequently work with raster data that includes temporal dimensions (e.g., PRISM, Daymet), you may find the `stars` package useful (covered in @sec-stars-basics). It offers a data model tailored for time-based raster data and allows the use of `dplyr` verbs for data manipulation.

### Direction for replication {-}

**Datasets**

All the datasets that you need to import are available [here](https://www.dropbox.com/sh/yf1u2gcnjyfbw38/AAD-cYgMyGMIP2kih2Jd6rjGa?dl=0). In this chapter, the path to files is set relative to my own working directory (which is hidden). To run the codes without having to mess with paths to the files, follow these steps:

+ set a folder (any folder) as the working directory using `setwd()`  
+ create a folder called "Data" inside the folder designated as the working directory (if you have created a "Data" folder previously, skip this step)
+ download the pertinent datasets from [here](https://www.dropbox.com/sh/yf1u2gcnjyfbw38/AAD-cYgMyGMIP2kih2Jd6rjGa?dl=0) 
+ place all the files in the downloaded folder in the "Data" folder

**Packages**

Run the following code to install or load (if already installed) the `pacman` package, and then install or load (if already installed) the listed package inside the `pacman::p_load()` function.

```{r Chap4_packages}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  terra, # handle raster data
  raster, # handle raster data
  mapview, # create interactive maps
  dplyr, # data wrangling
  sf, # vector data handling
  lubridate # date handling
)
```

```{r, echo = FALSE, eval = FALSE}
library(CropScapeR)

IA_cdl_2015 <-
  CropScapeR::GetCDLData(
    aoi = 19,
    year = "2015",
    type = "f"
  )

IA_cdl_2016 <-
  CropScapeR::GetCDLData(
    aoi = 19,
    year = "2016",
    type = "f"
  )

terra::writeRaster(IA_cdl_2015, here("Data/IA_cdl_2015.tif"), overwrite = TRUE)
terra::writeRaster(IA_cdl_2016, here("Data/IA_cdl_2016.tif"), overwrite = TRUE)
```

## Raster data object classes

### `raster` package: `RasterLayer`, `RasterStack`, and `RasterBrick`

Let's start with taking a look at raster data. We will use the CDL data for Iowa in 2015. We can use `raster::raster()` to read a raster data file.

```{r read_the_IA_cdl_data_run}
#| cache: false
(
  IA_cdl_2015 <- raster::raster("Data/IA_cdl_2015.tif")
)
```

Evaluating an imported raster object provides key information about the raster data, such as its dimensions (number of cells, rows, and columns), spatial resolution (e.g., 30 meters by 30 meters for this dataset), extent, coordinate reference system (CRS), and the minimum and maximum values recorded. The downloaded data is of the class `RasterLayer`, which is defined by the raster package. A `RasterLayer` contains only one layer, meaning that the raster cells hold the value of a single variable (in this case, the land use category code as an integer)

---

You can stack multiple raster layers of the **same spatial resolution and extent** to create a `RasterStack` using `raster::stack()` or `RasterBrick` using `raster::brick()`. Often times, processing a multi-layer object has computational advantages over processing multiple single-layer one by one^[You will see this in @sec-int-RV where we learn how to extract values from a raster layer for a vector data.]. 

To create a RasterStack and RasterBrick, let's load the CDL data for IA in 2016 and stack it with the 2015 data.

```{r make_stack_run}
#| cache: false

IA_cdl_2016 <- raster::raster("Data/IA_cdl_2016.tif")

#--- stack the two ---#
(
  IA_cdl_stack <- raster::stack(IA_cdl_2015, IA_cdl_2016)
)
```

`IA_cdl_stack` is of class `RasterStack`, and it has two layers of variables: CDL for 2015 and 2016. You can make it a `RasterBrick` using `raster::brick()`:

```{r make-brick, eval = F}
#--- stack the two ---#
IA_cdl_brick <- brick(IA_cdl_stack)

#--- or this works as well ---#
# IA_cdl_brick <- brick(IA_cdl_2015, IA_cdl_2016)

#--- take a look ---#
IA_cdl_brick
```

```{r, echo = F}
# saveRDS(IA_cdl_brick, "Data/IA_cdl_brick.rds")
(
IA_cdl_brick <- readRDS("Data/IA_cdl_brick.rds")
)
```

You probably noticed that it took some time to create the `RasterBrick` object^[Read [here](https://geocompr.robinlovelace.net/spatial-class.html#raster-classes) for the difference between `RasterStack` and `RasterBrick`]. While spatial operations on `RasterBrick` are supposedly faster than `RasterStack`, the time to create a `RasterBrick` object itself is often long enough to kill the speed advantage entirely. Often, the three raster object types are collectively referred to as `Raster`$^*$ objects for shorthand in the documentation of the `raster` and other related packages.

### `terra` package: `SpatRaster`

`terra` package has only one object class for raster data, `SpatRaster` and no distinctions between one-layer and multi-layer rasters is necessary. Let's first convert a `RasterLayer` to a `SpatRaster` using `terra::rast()` function.

```{r spat_raster, dependson = "read_the_IA_cdl_data", cache = F}
#--- convert to a SpatRaster ---#
IA_cdl_2015_sr <- terra::rast(IA_cdl_2015)

#--- take a look ---#
IA_cdl_2015_sr
```

You can see that the number of layers (`nlyr` in dimensions) is $1$ because the original object is a `RasterLayer`, which by definition has only one layer. Now, let's convert a `RasterStack` to a `SpatRaster` using `terra::rast()`.  

```{r spat_raster_nl, cache = F}
#--- convert to a SpatRaster ---#
IA_cdl_stack_sr <- terra::rast(IA_cdl_stack)

#--- take a look ---#
IA_cdl_stack_sr
```

Again, it is a `SpatRaster`, and you now see that the number of layers is 2. We just confirmed that `terra` has only one class for raster data whether it is single-layer or multiple-layer ones.

In order to make multi-layer `SpatRaster` from multiple single-layer `SpatRaster` you can just use `c()` like below:

```{r echo = F, cache = F}
# Keep getting this error below in SR-concatenate, so a work around:
# Error in x@ptr$combineSources(i@ptr) : external pointer is not valid
IA_cdl_2015_sr <- rast("Data/IA_cdl_2015.tif")
IA_cdl_2016_sr <- rast("Data/IA_cdl_2016.tif")
```

```{r SR-concatenate, cache = F}
#--- create a single-layer SpatRaster ---#
IA_cdl_2016_sr <- terra::rast(IA_cdl_2016)

#--- concatenate ---#
(
  IA_cdl_ml_sr <- c(IA_cdl_2015_sr, IA_cdl_2016_sr)
)
```

### Converting a `SpatRaster` object to a `Raster`$^*$ object.

You can convert a `SpatRaster` object to a `Raster`$^*$ object using `raster::raster()`, `raster::stack()`, and `raster::brick()`. Keep in mind that if you use `raster::rater()` even though `SpatRaster` has multiple layers, the resulting `RasterLayer` object has only the first of the multiple layers. 

```{r convert_back}
#--- RasterLayer (only 1st layer) ---#
IA_cdl_stack_sr %>% raster::raster()

#--- RasterLayer ---#
IA_cdl_stack_sr %>% raster::stack()

#--- RasterLayer (this takes some time) ---#
IA_cdl_stack_sr %>% raster::brick()
```

Instead of these functions, you can simply use `as(SpatRast, "Raster")` like below:

```{r }
as(IA_cdl_stack_sr, "Raster")
```

This works for any `Raster`$^*$ object and you do not have to pick the right function like above.

### Vector data in the `terra` package

`terra` package has its own class for vector data, called `SpatVector`. While we do not use any of the vector data functionality provided by the `terra` package, we learn how to convert an `sf` object to `SpatVector` because some of the `terra` functions do not support `sf` as of now (this will likely be resolved very soon). We will see some use cases of this conversion in @sec-int-RV when we learn raster value extractions for vector data using `terra::extract()`. 

As an example, let's use Illinois county border data. 

```{r il_county_echo}
#| cache: true
#--- Illinois county boundary ---#
(
  IL_county <- 
    tigris::counties(
      state = "Illinois", 
      progress_bar = FALSE
    ) %>%
    dplyr::select(STATEFP, COUNTYFP)
)
```

You can convert an `sf` object to `SpatVector` object using `terra::vect()`.

```{r to_sv}
(
  IL_county_sv <- terra::vect(IL_county)
)
```

## Read and write a raster data file

Raster data files can come in numerous different formats. For example, PRPISM comes in the Band Interleaved by Line (BIL) format, some of the Daymet data comes in netCDF format. Other popular formats include GeoTiff, SAGA, ENVI, and many others. 

### Read raster file(s)

You can use `terra::rast()` to read raster data in many common formats, and in most cases, this function will work for your raster data. In this example, we read a GeoTIFF file (with a .tif extension).

```{r read_no_eval_terra, cache  =F}
(
  IA_cdl_2015_sr <- terra::rast("Data/IA_cdl_2015.tif")
)
```

You can read multiple single-layer raster datasets of the same spatial extent and resolution at the same time to have a multi-layer `SpatRaster` object. Here, we import two single-layer raster datasets (IA_cdl_2015.tif and IA_cdl_2016.tif) to create a two-layer `SpatRaster` object.

```{r multiple_files}
#--- the list of path to the files ---#
files_list <- c("Data/IA_cdl_2015.tif", "Data/IA_cdl_2016.tif")

#--- read the two at the same time ---#
(
  multi_layer_sr <- terra::rast(files_list)
)
```

Of course, this only works because the two datasets have the identical spatial extent and resolution. There are, however, no restrictions on what variable each of the raster layers represent. For example, you can combine PRISM temperature and precipitation raster layers if you want.

### Write raster files

You can write a `SpatRaster` object using `terra::writeRaster()`.

```{r write_terra, eval = F}
terra::writeRaster(IA_cdl_2015_sr, "Data/IA_cdl_stack.tif", filetype = "GTiff", overwrite = TRUE)
```

The above code saves `IA_cdl_2015_sr` (a `SpatRaster` object) as a GeoTiff file.^[There are many other alternative formats (see [here](https://www.rdocumentation.org/packages/raster/versions/3.0-12/topics/writeRaster))] The filetype option can be dropped as `writeRaster()` infers the filetype from the extension of the file name. The `overwrite = TRUE` option is necessary if a file with the same name already exists and you are overwriting it. This is one of the many areas `terra` is better than `raster`. `raster::writeRaster()` can be frustratingly slow for a large `Raster`$^*$ object. `terra::writeRaster()` is much faster.

You can also save a multi-layer `SpatRaster` object just like you save a single-layer `SpatRaster` object. 

```{r write_terra_2, eval = F}
terra::writeRaster(IA_cdl_stack_sr, "Data/IA_cdl_stack.tif", filetype = "GTiff", overwrite = TRUE)
```

The saved file is a multi-band raster datasets. So, if you have many raster files of the same spatial extent and resolution, you can "stack" them on R and then export it to a single multi-band raster datasets, which cleans up your data folder.
 
## Extract information from raster data object

### Get CRS

You often need to extract the CRS of a raster object before you interact it with vector data (e.g., extracting values from a raster layer to vector data, or cropping a raster layer to the spatial extent of vector data), which can be done using `terra::crs()`: 

```{r get_crs_terra}
terra::crs(IA_cdl_2015_sr)
```

### Subset

You can access specific layers in a multi-layer raster object by indexing:

```{r }
#--- index ---#
IA_cdl_stack_sr[[2]] # (originally IA_cdl_2016.tif)
```

### Get cell values

You can access the values stored in a `SpatRaster` object using the `terra::values()` function:

```{r values, eval = F}
#--- terra::values ---#
values_from_rs <- terra::values(IA_cdl_stack_sr)

#--- take a look ---#
head(values_from_rs)
``` 

```{r values-show, echo = F}
# saveRDS(head(values_from_rs), "Data/values-from-rs.rds")

(
  values_from_rs <- readRDS("Data/values-from-rs.rds")
)
``` 

The returned values come in a matrix form of two columns because we are getting values from a two-layer `SpatRaster` object (one column for each layer). In general, `terra::values()` returns a $X$ by $n$ matrix, where $X$ is the number of cells and $n$ is the number of layers. 

## Turning a raster object into a `data.frame` (not necessary)

You can use the `as.data.frame()` function with `xy = TRUE` option to construct a `data.frame` where each row represents a single cell that has cell values for each layer and its coordinates (the center of the cell).

```{r to-dataframe, eval = F}
#--- converting to a data.frame ---#
IA_cdl_df <- as.data.frame(IA_cdl_stack_sr, xy = TRUE) # this works with Raster* objects as well

#--- take a look ---#
head(IA_cdl_df)
```

```{r to-dataframe-show, echo = F}
# saveRDS(head(IA_cdl_df), "Data/IA_CDL_df.rds")

readRDS("Data/IA_CDL_df.rds")
```

:::{.callout-important title="Caveat"}
I have seen cases where `raster` objects are converted to a `data.frame` and then to an `sf` object for interacting with polygons using `sf::st_join()` to extract and assign cell values to polygons (see @sec-int-vv for this type of operation). However, this approach is generally not recommended for two main reasons.

First, it is significantly slower than using functions designed to work directly with raster objects and polygons, such as `terra::extract()` or `exactextractr::exact_extract()`, which are introduced in @sec-int-RV. The primary reason is that converting a raster to a data.frame is a time-consuming process.

Second, once a raster object is converted to point `sf` data, it becomes impossible to weight cell values based on their degree of overlap with the target polygons. While working with a `data.frame` may be appealing due to their familiarity, the conversion is often unnecessary and inefficient^[If you know of cases where converting to a `data.frame` is beneficial, please let me know, and I will include them here.].
:::

## Quick visualization

To have a quick visualization of the data values of `SpatRaster` objects, you can simply use `plot()`:

```{r plot_stack, cache = TRUE}
plot(IA_cdl_2015_sr)
```

For a more elaborate map using raster data, see @sec-geom-raster. 


## Working with netCDFs {#sec-work-with-netcdf}

A netCDF file contains data with a specific structure: a two-dimensional spatial grid (e.g., longitude and latitude) and a third dimension which is usually date or time. This structure is convenient for weather data measured on a consistent grid over time. One such dataset is called [gridMET](http://www.climatologylab.org/gridmet.html) which maintains a gridded dataset of weather variables at 4km resolution. Let's download the daily precipitation data for 2018 using `downloader::download()`^[gridMET data is also available in the [Google Earth Engine Data Catalog](https://developers.google.com/earth-engine/datasets/), which can be accessed with the R library [`rgee`](https://github.com/r-spatial/rgee)]. We set the destination file name (what to call the file and where we want it to be), and the mode to `wb` for a binary download. 

```{r eval = F}
#--- download gridMET precipitation 2018 ---#
downloader::download(
  url = str_c("http://www.northwestknowledge.net/metdata/data/pr_2018.nc"),
  destfile = "Data/pr_2018.nc",
  mode = "wb"
)
```

This code should have stored the data as **pr_2018.nc** in the **Data** folder. You can read a netCDF file using `terra::rast()`.

```{r }
(
  pr_2018_gm <- terra::rast("Data/pr_2018.nc")
)
```

You can see that it has 365 layers: one layer per day in 2018. Let's now look at layer names:  

```{r }
head(names(pr_2018_gm))
```

Since we have 365 layers and the number at the end of the layer names increase by 1, you would think that **n**th layer represents **n**th day of 2018. In this case, you are correct. However, it is always a good practice to confirm what each layer represents without assuming anything. Now, let's use the `ncdf4` package, which is built specifically to handle netCDF4 objects.

```{r }
(
  pr_2018_nc <- ncdf4::nc_open("Data/pr_2018.nc")
)
```

As you can see from the output, there is tons of information that we did not see when we read the data using `rast()`, which includes the explanation of the third dimension (day) of this raster object. It turned out that the numerical values at the end of layer names in the `SpatRaster` object are **days since 1900-01-01**. So, the first layer (named **precipitation_amount_day=43099**) represents:

```{r }
lubridate::ymd("1900-01-01") + 43099
```

Actually, if we use `raster::brick()`, instead of `terra::rast()`, then we can see the naming convention of the layers:

```{r }
(
  pr_2018_b <- raster::brick("Data/pr_2018.nc")
)
```

`SpatRaster` or `RasterBrick` objects are easier to work with as many useful functions accept them as inputs, but not the `ncdf4` object. Personally, I first scrutinize a netCDFs file using `nc_open()` and then import it as a `SpatRaster` or `RasterBrick` object^[Even though `RasterBrick` provides the description of how layers are named, I think it is a good practice to see the full description in the `ncdf4` object.]. Recovering the dates for the layers is particularly important as we often wrangle the resulting data based on date (e.g., subset the data so that you have only April to September). An example of date recovery can be seen in @sec-gridMET.

For those who are interested in more detailed descriptions of how to work with `ncdf4` object is provided [here](https://pjbartlein.github.io/REarthSysSci/netCDF.html).


