### Project Overview

---

**Objective**

+ Understand the impact of past precipitation on crop choice in Iowa (IA).

---

**Datasets**

+ IA county boundary 
+ Regular grids over IA, created using `sf::st_make_grid()` 
+ PRISM daily precipitation data downloaded using `prism` package
+ Land use data from the Cropland Data Layer (CDL) for IA in 2015, downloaded using `CropScapeR` package

---

**Econometric Model**

The econometric model we would like to estimate is:

$$
CS_i = \alpha + \beta_1 PrN_{i} + \beta_2 PrC_{i} + v_i
$$

where $CS_i$ is the area share of corn divided by that of soy in 2015 for grid $i$ (we will generate regularly-sized grids in the Demo section), $PrN_i$ is the total precipitation observed in April through May and September  in 2014, $PrC_i$ is the total precipitation observed in June through August in 2014, and $v_i$ is the error term. To run the econometric model, we need to find crop share and weather variables observed at the grids. We first tackle the crop share variable, and then the precipitation variable.

---

**GIS tasks**

+ download Cropland Data Layer (CDL) data by USDA NASS 
  * use `CropScapeR::GetCDLData()`
+ download PRISM weather data
  * use `prism::get_prism_dailys()`
+ crop PRISM data to the geographic extent of IA 
  * use `terra::crop()`
* read PRISM data
  * use `terra::rast()`
* extract the CRS of PRISM data
  * use `terra::crs()`
+ create regular grids within IA, which become the observation units of the econometric analysis
  * use `sf::st_make_grid()` 
+ remove grids that share small area with IA 
  * use `sf::st_intersection()` and `sf::st_area`
+ assign crop share and weather data to each of the generated IA grids (parallelized)
  * use `exactextractr::exact_extract()` and `future.apply::future_lapply()`
+ create maps 
  * use the `tmap` package 

---

**Preparation for replication**

+ Run the following code to install or load (if already installed) the `pacman` package, and then install or load (if already installed) the listed package inside the `pacman::p_load()` function.

```{r demo3_packages}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  sf, # vector data operations
  terra, # raster data operations
  exactextractr, # fast raster data extraction for polygons
  maps, # to get county boundary data
  data.table, # data wrangling
  dplyr, # data wrangling
  lubridate, # Date object handling
  tmap, # for map creation
  modelsummary, # regression table generation
  future.apply, # parallel computation
  CropScapeR, # download CDL data
  prism, # download PRISM data
  stringr # string manipulation
)
```

### Project Demonstration

The geographic focus of this project is Iowas. Let's get Iowa state border (see @fig-IA-map for its map).

```{r Demo-3-IA-boundary}
#--- IA state boundary ---#
IA_boundary <- sf::st_as_sf(tigris::counties(state = "Iowa", cb = TRUE, progress_bar = FALSE))
```

::: {.column-margin}
```{r}
#| label: fig-IA-map
#| fig-cap: "Iowa state boundary"
#| code-fold: true
#--- map IA state border ---#
tm_shape(IA_boundary) +
  tm_polygons() +
  tm_layout(frame = FALSE)
```
:::
<!--end of column-margin-->

The unit of analysis is artificial grids that we create over Iowa. The grids are regularly-sized rectangles except around the edge of the Iowa state border^[We by no means are saying that this is the right geographical unit of analysis. This is just about demonstrating how R can be used for analysis done at the higher spatial resolution than county.]. So, let's create grids and remove those that do not overlap much with Iowa (@fig-Demo-3-IA-grids-map shows what the generated grids look like).

```{r Demo-3-IA-grids} 
#--- create regular grids (40 cells by 40 columns) over IA ---#
IA_grids <-
  IA_boundary %>%
  #--- create grids ---#
  sf::st_make_grid(, n = c(40, 40)) %>%
  #--- convert to sf ---#
  sf::st_as_sf() %>%
  #--- assign grid id for future merge ---#
  dplyr::mutate(grid_id = 1:nrow(.)) %>%
  #--- make some of the invalid polygons valid ---#
  sf::st_make_valid() %>%
  #--- drop grids that do not overlap with Iowa ---#
  .[IA_boundary, ]
```

::: {.column-margin}
```{r}
#| label: fig-Demo-3-IA-grids-map
#| fig-cap: "Map of regular grids generated over IA"
#| code-fold: true
#--- plot the grids over the IA state border ---#
tm_shape(IA_boundary) +
  tm_borders(col = "red", lwd = 2) +
  tm_shape(IA_grids) +
  tm_borders(col = "blue") +
  tm_layout(frame = FALSE)
```
:::
<!--end of column-margin-->

---

Let's work on crop share data. You can download CDL data using the `CropScapeR::GetCDLData()` function.

```{r IA_data_import, echo = F}
IA_cdl_2015 <- terra::rast("Data/IA_cdl_2015.tif")
```

```{r Demo-3_cdl_download, eval = F}
#--- download the CDL data for IA in 2015 ---#
(
  cdl_IA_2015 <- CropScapeR::GetCDLData(aoi = 19, year = 2015, type = "f")
)
```

The cells (30 meter by 30 meter) of the imported raster layer take a value ranging from 0 to 255. Corn and soybean are represented by 1 and 5, respectively.

@fig-overlap-cdl-grid shows the map of one of the IA grids and the CDL cells it overlaps with.

```{r}
#| label: fig-overlap-cdl-grid
#| fig-cap: "Spatial overlap of an IA grid and CDL layer"
#| code-fold: true
temp_grid <- IA_grids[100, ]

extent_grid <-
  temp_grid %>%
  sf::st_transform(., terra::crs(IA_cdl_2015)) %>%
  sf::st_bbox()

raster_ovelap <- terra::crop(IA_cdl_2015, extent_grid)

ggplot() +
  tidyterra::geom_spatraster(data = raster_ovelap, aes(fill = Layer_1)) +
  geom_sf(data = temp_grid, fill = NA, color = "red", linewidth = 1) +
  scale_fill_viridis_c() +
  theme_void()
```

We would like to extract all the cell values within the red border.

We use `exactextractr::exact_extract()` to identify which cells of the CDL raster layer fall within each of the IA grids and extract land use type values. We then find the share of corn and soybean for each of the grids.

```{r Demo-3_extract}
#| eval: false
#| results: "hide"
#--- reproject grids to the CRS of the CDL data ---#
IA_grids_rp_cdl <- sf::st_transform(IA_grids, terra::crs(IA_cdl_2015))

#--- extract crop type values and find frequencies ---#
cdl_extracted <-
  exactextractr::exact_extract(IA_cdl_2015, IA_grids_rp_cdl) %>%
  lapply(., function(x) data.table(x)[, .N, by = value]) %>%
  #--- combine the list of data.tables into one data.table ---#
  data.table::rbindlist(idcol = TRUE) %>%
  #--- find the share of each land use type ---#
  .[, share := N / sum(N), by = .id] %>%
  .[, N := NULL] %>%
  #--- keep only the share of corn and soy ---#
  .[value %in% c(1, 5), ]
```

```{r, echo = F}
IA_grids_rp_cdl <- sf::st_transform(IA_grids, terra::crs(IA_cdl_2015))

# saveRDS(cdl_extracted, "Data/cdl_extracted.rds")
cdl_extracted <- readRDS("Data/cdl_extracted.rds")
```

We then find the corn to soy ratio for each of the IA grids.

```{r Demo-3_share_calc}
#--- find corn/soy ratio ---#
corn_soy <-
  cdl_extracted %>%
  #--- long to wide ---#
  data.table::dcast(.id ~ value, value.var = "share") %>%
  #--- change variable names ---#
  data.table::setnames(c(".id", "1", "5"), c("grid_id", "corn_share", "soy_share")) %>%
  #--- corn share divided by soy share ---#
  .[, c_s_ratio := corn_share / soy_share]
```

---

We are still missing daily precipitation data at the moment. We have decided to use daily weather data from PRISM. Daily PRISM data is a raster data with the cell size of 4 km by 4 km. @fig-Demo-3-show-prism-data presents precipitation data downloaded for April 1, 2010. It covers the entire contiguous U.S.

```{r}
#| label: fig-Demo-3-show-prism-data
#| fig-cap: "Map of PRISM raster data layer"   
#| code-fold: true
prism_ex <- readRDS("Data/prism_ex.rds")

plot(prism_ex)
```

Let's now download PRISM data (You do not have to run this code to get the data. It is included in the data folder for replication [here](https://www.dropbox.com/sh/rtbs4ji21c9uiy9/AADYpHAWhUxMittAptuq-Apaa?dl=0)). This can be done using the `get_prism_dailys()` function from the `prism` package.^[[prism Github page](https://github.com/ropensci/prism)]  

<!-- not to be seen -->
```{r Demo-3_get_prism, eval = FALSE}
options(prism.path = "Data/PRISM")

prism::get_prism_dailys(
  type = "ppt",
  minDate = "2014-04-01",
  maxDate = "2014-09-30",
  keepZip = FALSE
)
```

When we use `get_prism_dailys()` to download data^[For this project, monthly PRISM data could have been used, which can be downloaded using the `prism::get_prism_monthlys()` function. But, in many applications, daily data is necessary, so how to download and process them is illustrated here.], it creates one folder for each day. So, I have about 180 folders inside the folder I designated as the download destination above with the `options()` function. 

<!-- The name of the folder is expressive about what the data inside it is about. For example, the precipitation data for April 1st, 2010 is stored in the folder called "PRISM_ppt_stable_4kmD2_20100401_bil." Inside it, you will see bunch of files with exactly the same prefix, but with different extensions.   --> 

---

We now try to extract precipitation value by day for each of the IA grids by geographically overlaying IA grids onto the PRISM data layer and identify which PRISM cells each of the IA grid encompass. ^[Be cautious about using `sf::st_buffer()` for spatial objects in geographic coordinates (latitude, longitude) in practice. Significant distortion will be introduced to the buffer due to the fact that one degree in latitude and longitude means different distances at the latitude of IA. Here, I am just creating a buffer to extract PRISM cells to display on the map.]. 

```{r}
#--- read a PRISM dataset ---#
prism_whole <- terra::rast("Data/PRISM/PRISM_ppt_stable_4kmD2_20140401_bil/PRISM_ppt_stable_4kmD2_20140401_bil.bil")

#--- align the CRS ---#
IA_grids_rp_prism <- sf::st_transform(IA_grids, terra::crs(prism_whole))

#--- crop the PRISM data for the 1st IA grid ---#
sf_use_s2(FALSE)
PRISM_1 <- terra::crop(prism_whole, sf::st_buffer(IA_grids_rp_prism[1, ], dist = 0.05))
```

@fig-Demo-3-prism-crop shows how the first IA grid (in red) overlaps with the PRISM cells. As you can see, some PRISM grids are fully inside the analysis grid, while others are partially inside it. So, when assigning precipitation values to grids, we will use the coverage-weighted mean of precipitations.

```{r}
#| label: fig-Demo-3-prism-crop
#| fig-cap: "Spatial overlap of an IA grid over PRISM cells"

#--- map them ---#
ggplot() +
  tidyterra::geom_spatraster(data = PRISM_1) +
  scale_fill_viridis_c(name = "Precipitation") +
  geom_sf(data = IA_grids_rp_prism[1, ], fill = NA, color = "red") +
  theme_void()
```

Unlike the CDL layer, we have `r seq(as.Date("2014-04-01"), as.Date("2014-09-30"), "days") %>% length` raster layers to process. Fortunately, we can process many raster files at the same time very quickly by first "stacking" many raster files first and then applying the `exactextractr::exact_extract()` function. Using `future.apply::future_lapply()`, we let $6$ cores take care of this task with each processing 31 files, except one of them handling only 28 files.^[Parallelization of extracting values from many raster layers for polygons are discussed in much more detail in @sec-EE.]

We first get all the paths to the PRISM files. 

```{r setup_parallel}
#--- get all the dates ---#
dates_ls <- seq(as.Date("2014-04-01"), as.Date("2014-09-30"), "days")

#--- remove hyphen ---#
dates_ls_no_hyphen <- stringr::str_remove_all(dates_ls, "-")

#--- get all the prism file names ---#
folder_name <- paste0("PRISM_ppt_stable_4kmD2_", dates_ls_no_hyphen, "_bil")
file_name <- paste0("PRISM_ppt_stable_4kmD2_", dates_ls_no_hyphen, "_bil.bil")
file_paths <- paste0("Data/PRISM/", folder_name, "/", file_name)

#--- take a look ---#
head(file_paths)
```

We now prepare for parallelized extractions and then implement them using `future_apply()` (you can have a look at @sec-par-comp to familiarize yourself with parallel computation using the `future.apply` package).

```{r go_parallel_prep}
#--- define the number of cores to use ---#
num_core <- 6

#--- prepare some parameters for parallelization ---#
file_len <- length(file_paths)
files_per_core <- ceiling(file_len / num_core)

#--- prepare for parallel processing ---#
future::plan(multicore, workers = num_core)

#--- reproject IA grids to the CRS of PRISM data ---#
IA_grids_reprojected <- sf::st_transform(IA_grids, terra::crs(prism_whole))
```

Here is the function that we run in parallel over `r num_core` cores. 

```{r define_function_prism_get, eval = F}
#--- define the function to extract PRISM values by block of files ---#
extract_by_block <- function(i, files_per_core) {

  #--- files processed by core  ---#
  start_file_index <- (i - 1) * files_per_core + 1

  #--- indexes for files to process ---#
  file_index <- seq(
    from = start_file_index,
    to = min((start_file_index + files_per_core), file_len),
    by = 1
  )

  #--- extract values ---#
  data_temp <- 
    file_paths[file_index] %>% # get file names
    #--- read as a multi-layer raster ---#
    terra::rast() %>%
    #--- extract ---#
    exactextractr::exact_extract(., IA_grids_reprojected) %>%
    #--- combine into one data set ---#
    data.table::rbindlist(idcol = "ID") %>%
    #--- wide to long ---#
    data.table::melt(id.var = c("ID", "coverage_fraction")) %>%
    #--- calculate "area"-weighted mean ---#
    .[, .(value = sum(value * coverage_fraction) / sum(coverage_fraction)), by = .(ID, variable)]

  return(data_temp)
}
```

Now, let's run the function in parallel and calculate precipitation by period.

```{r parallel_prism_not_run, eval = FALSE}
#--- run the function ---#
precip_by_period <-
  future.apply::future_lapply(
    1:num_core,
    function(x) extract_by_block(x, files_per_core)
  ) %>%
  data.table::rbindlist() %>%
  #--- recover the date ---#
  .[, variable := as.Date(str_extract(variable, "[0-9]{8}"), "%Y%m%d")] %>%
  #--- change the variable name to date ---#
  data.table::setnames("variable", "date") %>%
  #--- define critical period ---#
  .[, critical := "non_critical"] %>%
  .[month(date) %in% 6:8, critical := "critical"] %>%
  #--- total precipitation by critical dummy  ---#
  .[, .(precip = sum(value)), by = .(ID, critical)] %>%
  #--- wide to long ---#
  data.table::dcast(ID ~ critical, value.var = "precip")
```

```{r read_precip_period, echo = FALSE}
# saveRDS(precip_by_period, "Data/precip_by_period.rds")
precip_by_period <- readRDS("Data/precip_by_period.rds")
```

We now have grid-level crop share and precipitation data. 

---

Let's merge them and run regression.^[We can match on `grid_id` from `corn_soy` and `ID` from "precip_by_period" because `grid_id` is identical with the row number and ID variables were created so that the ID value of $i$ corresponds to $i$ th row of `IA_grids`.]

```{r Demo-3-reg}
#--- crop share ---#
reg_data <- corn_soy[precip_by_period, on = c(grid_id = "ID")]

#--- OLS ---#
reg_results <- lm(c_s_ratio ~ critical + non_critical, data = reg_data)
```

Here is the regression results table.


```{r Demo-3-reg-table}
#--- regression table ---#
modelsummary::msummary(
  reg_results,
  stars = TRUE,
  gof_omit = "IC|Log|Adj|Within|Pseudo"
)
```

Again, do not read into the results as the econometric model is terrible.


